---
title: "Vignette Writeup"
author: "Edward Ho, "
date: last-modified
published-title: "Updated"
editor: visual
format: html
code-copy: true
execute:
  message: false
  warning: false
  echo: false
  cache: true
---

```{r}
# load any other packages and read data here
library(tidyverse)
library(ggplot2)
library(gridExtra)
library(knitr)
```

# Executive summary

Time series anomaly detection and change-point on the univariate (potentially multivariate case) for time series economic data from LA concerning unemployment.

## Task Question 1:

[TEXT HERE]

## Task Question 2:

The primary interest in task 2 was to find significant periods of changes, whereas task 1 focused on anomaly detection of certain points, our interest lies in finding significant increments of time where there may have been upwards or downwards shifts in unemployment rates.

# Data description

```{r}
unemployment <- read_csv('data/processed_data.csv')
unemployment
```

Above we can see that the data collected is essentially a wrap-up on the 1st of a given month from the years 1900 to 2021 recorded from the city of Los Angeles. These variables encapsulate various city markers that range from unemployment to government benefits. 
Notably, belows are some interesting variables of interest:

- The date (Year-Month-Day) of each observation/recording

- The unemployment rate

- Average price of electricity

- Average price of gasoline 

Since we are only concerned with the time-series aspect of unemployment, we are only focusing on the date and the unemployment rate of each month, therefore we will not show much interest in the other variables, yet.

# Findings

[Finding's and Methodology for Task 1 and 2 below]

## Task 1

[things of interest]

### Methodology for Task 1

[how it's found]

### Results

[finding's]

## Task 2

[things of interest]

### Methodology for Task 2

[how it's found]

### Results

[finding's]

# Conclusion

[wrapping up project, final remark's about project/data/results]